For induction-induction, induction is enough

### 2019-09-25 paper/universaliit.tex
Subsections on recursion and uniqueness of the initial morphism

  \todo[inline]{András: Hugunin's paper has an example (which is
     likely generalizable) that a notion of redundancy in
     well-formedness relations implies that UIP is required.}
  \subsubsection{Stability under weakening}
  Let us argue that it is necessary to show stability of the relation under
  weakening before showing that it is stable under substitution.
  Indeed, in the proof of stability under substitution, the $\Pi$ case
  requires to show that $\Pi \,A[\sigma] \,B[\keep\,\sigma]$ is related to
  $\model{\Pi}\,\model{A}\model{[\sigma]}
  \model{A}\model{[\model{\keep}\,\sigma]}$.
  We would like to apply the induction hypothesis, so we need to show that
  $\keep\,\sigma=\varu\,0\consu\wkO\,\sigma$ is related to
  $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
  related to $\model{\sigma}$.
  As $\keep\,\sigma=\varu 0\consu \wkO\,\sigma$, we are left with showing that
  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO \id$)
  relates to its semantic counterpart.

@@

\subsection{Existence and uniqueness of the initial QIIT morphism}
Induction on the typing judgments shows that any QIIT morphism from the
syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
semantic counter-part. More formally, we show by recursion:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  Right uniqueness of the relation is used in this recursion.

It is then straightforward to show that the first projection of these
constructions yield a model morphism from the syntax to the model, using right
uniqueness of the relation.
Now, to show that it is unique, suppose that we have a morphism from the
syntax to the model, inducing the following maps:
\begin{alignat*}{5}
  &
  \mor{\Con}
  && :
   (\Gamma \vdash) \ra \model\Con
   \\
  &
  \mor{\Ty}
  && :
   (\wf\Gamma:\Gamma \vdash) \ra (\Gamma\vdash A)\ra\model\Ty\,(\mor\Con\wf\Gamma)
   \\
  &
  \mor{\Tm}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
  (\Gamma\vdash t \in A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
  (\mor\Ty\wf\Gamma\,\wf{A})
   \\
  &
  \mor{\Var}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
  (\Gamma\vdash x \invar A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
  (\mor\Ty\wf\Gamma\,\wf{A})
   \\
  &
  \mor{\Sub}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra
  (\wf\Delta:\Delta \vdash) \ra
  (\Gamma\vdash \sigma \Rightarrow \Delta)\ra\model\Sub\,(\mor\Con\wf\Gamma)\,
  (\mor\Con\wf\Delta)
\end{alignat*}
Then, we show by recursion on the typing judgments that
the image is related:
\begin{alignat*}{5}
  &
  {\sim}\mor{\Con}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra \wf\Gamma\sim \mor\Con\,\wf\Gamma
  \\
  &
  {\sim}\mor{\Ty}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  \wf\Gamma\sim \mor\Ty\,\wf\Gamma\,\wf{A}
  \\
  &
  {\sim}\mor{\Tm}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{t} : \Gamma \vdash t \in A) \ra
  \wf\Gamma\sim \mor\Tm\,\wf\Gamma\,\wf{A}\,\wf{t}
  \\
  &
  {\sim}\mor{\Var}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{x} : \Gamma \vdash x \invar A) \ra
  \wf\Gamma\sim \mor\Var\,\wf\Gamma\,\wf{A}\,\wf{x}
  \\
  &
  {\sim}\mor{\Sub}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{\Delta} : \Delta \vdash) \ra
  (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \wf\Gamma\sim \mor\Sub\,\wf\Gamma\,\wf{\Delta}\,\wf{\sigma}
\end{alignat*}
This justifies the uniqueness of the morphism, by right uniqueness of the relation.

\subsection{Recursion}
\subsection{Uniqueness}

>>>>

  % \todo[inline]{András: Hugunin's paper has an example (which is
  %    likely generalizable) that a notion of redundancy in
  %    well-formedness relations implies that UIP is required.}
  \subsubsection{Stability under weakening}
  % Let us argue that it is necessary to show s
  Stability of the relation under
  weakening must be proved before stability under substitution.
  Indeed, in the proof of stability under substitution, the $\Pi$ case
  requires to show that $\Pi \,A[\sigma] \,B[\keep\,\sigma]$ is related to
  $\model{\Pi}\,\model{A}\model{[\sigma]}
  \model{A}\model{[\model{\keep}\,\sigma]}$.
  We would like to apply the induction hypothesis, so we need to show that
  $\keep\,\sigma=\varu\,0\consu\wkO\,\sigma$ is related to
  $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
  related to $\model{\sigma}$.
  As $\keep\,\sigma=\varu 0\consu \wkO\,\sigma$, we are left with showing that
  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO \id$)
  relates to its semantic counterpart.

@@

\subsection{Recursion}
Recursion consists in building a morphism from the syntactic model to the
semantic one.
The image of a syntactic context is going to be a semantic context which relates
to it (which is unique thanks to right uniqueness of the relation), and
similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we construct by induction on well-typed judgments the
semantic counter-parts of the syntax:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  Right uniqueness of the relation is used in this induction.

It is then straightforward to show (without induction) that the first projection of these
constructions yield a model morphism from the syntax to the model, again using right
uniqueness of the relation.
\subsection{Uniqueness}
Our goal is to show that the syntactic model is initial. What remains thus is to
show that the morphism constructed in the previous section is actually the
unique model morphism between the syntax and the arbitray model.
To this end, we exploit right uniqueness of the relation: it is then enough to show
that any such morphism maps a syntactic context to a related semantic context,
and similarly for types and terms.

More formally, we assume an arbitrary morphism from the
syntax to the model, inducing the following maps:
\begin{alignat*}{5}
  &
  \mor{\Con}
  && :
   (\Gamma \vdash) \ra \model\Con
   \\
  &
  \mor{\Ty}
  && :
   (\wf\Gamma:\Gamma \vdash) \ra (\Gamma\vdash A)\ra\model\Ty\,(\mor\Con\wf\Gamma)
   \\
  &
  \mor{\Tm}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
  (\Gamma\vdash t \in A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
  (\mor\Ty\wf\Gamma\,\wf{A})
   \\
  &
  \mor{\Var}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
  (\Gamma\vdash x \invar A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
  (\mor\Ty\wf\Gamma\,\wf{A})
   \\
  &
  \mor{\Sub}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra
  (\wf\Delta:\Delta \vdash) \ra
  (\Gamma\vdash \sigma \Rightarrow \Delta)\ra\model\Sub\,(\mor\Con\wf\Gamma)\,
  (\mor\Con\wf\Delta)
\end{alignat*}
Then, we show by recursion on the typing judgments that
the image is related:
\begin{alignat*}{5}
  &
  {\sim}\mor{\Con}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra \wf\Gamma\sim \mor\Con\,\wf\Gamma
  \\
  &
  {\sim}\mor{\Ty}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  \wf\Gamma\sim \mor\Ty\,\wf\Gamma\,\wf{A}
  \\
  &
  {\sim}\mor{\Tm}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{t} : \Gamma \vdash t \in A) \ra
  \wf\Gamma\sim \mor\Tm\,\wf\Gamma\,\wf{A}\,\wf{t}
  \\
  &
  {\sim}\mor{\Var}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{x} : \Gamma \vdash x \invar A) \ra
  \wf\Gamma\sim \mor\Var\,\wf\Gamma\,\wf{A}\,\wf{x}
  \\
  &
  {\sim}\mor{\Sub}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{\Delta} : \Delta \vdash) \ra
  (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \wf\Gamma\sim \mor\Sub\,\wf\Gamma\,\wf{\Delta}\,\wf{\sigma}
\end{alignat*}
This justifies the uniqueness of the morphism, by right uniqueness of the relation.
